---
title: "R packages and how to use them"
---

So far we've only used functions supplied as part of base R. However the real power of R comes from the ability to import functions
the community of R programmers has built. These come in packages, which are a series of related functions bundled together

Generally speaking before implementing your function it is always worth checking if there is a well-maintained and documented package that performs what you would like to do.

# Installing packages

Most R packages belong to a repository known as CRAN. CRAN packages can be installed with  `install.packages("package_name")`

Often R will ask for your permission to install packages compiled from source rather than binary releases ^[but wait? what is compilation?.\n
unlike R some programming languages need to be translated from a human readable text format to a binary file understandable by the computer.\n R packages often interface with code in compiled languages as it is generally faster]. If you want to use the absolutely most recent version say yes, if you're in a hurry say no. Installing packages from source on Windows requires an additional component: [RTools](https://cran.r-project.org/bin/windows/Rtools/). 

# Using packages

packages can be loaded by using the library function.
Conventionally, packages are loaded at the beginning of a script, to make obvious which packages are needed.
```{r warning=TRUE}
library(ggplot2)
```

once loaded you can use the functions in the package as we have done so far
We can use sessionInfo() to display information on R and the loaded packages. 
```{r}
sessionInfo()
```

## Documentation 

all packages have more or less extensive documentation. A quick Google search of your package and R or CRAN should bring up a documentation page similar to the one for [cowplot](https://cran.r-project.org/web/packages/cowplot/index.html). 
All packages have a reference manual, detailing one by one the use of all the functions. This is the same information that the help panel uses so it can be quite dry.  

Additionally most packages have a vignette, introducing the package and showing how to actually use it to solve a problem.

we can also view the documentation from R with:
```{r eval=FALSE}
browseVignettes("tidyverse") # shows the vignette
?cowplot # shows the reference manual
```

## citing packages

Package authors, particularly in bioinformatics are often Academics. If you use a package consider citing any paper in which they presented their work ^[Of course, packages are often built on top of other packages, in most cases it would not be reasonable to cite all the packages used and all their dependencies.].
We can use the `citation("package")` function to retrieve the citation information.

for instance the IRanges package suggests you cite a PLOS Computational Biology
```{r}
citation("IRanges")
```
# key packages

This section will introduce some key packages,

## Tidyverse

the tidyverse is a collection of closely related packages for Data Science.
They can be installed and loaded as one with `install.packages("tidyverse").
The 

The tidyverse packages provide functions to clean, manipulate, analyse and visualise data.

In tidy data:
 
1. Every column is a variable.
2. Every row is an observation.
3. Every cell is a single value.

you can read more about tidy data [here](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html)

The tidyverse is composed of:

- ggplot2
- tibble
- tidyr
- readr
- purr
- dyplr
- stringr
- forcats

## ggplot2 

ggplot2 is a plotting package. while base R plots can become very beautiful with lots of customisation,
ggplot2 can easily produces publication quality graphics with comparatively little effort

GGplot is based on the grammar of graphics, a whole theory of how to describe a plot whose explanation is 
best left to the [official book](https://ggplot2-book.org/index.html). In short a ggplot2 graph is composed
of 3 main elements: data, a geometry representing the individual data points and a mapping

for a gallery of what is possible with ggplot and other plotting packages see 
[r graph gallery ](https://www.r-graph-gallery.com/histogram)
ggplot2's grammar can appear complex at first, so we'll cover some common plots in session 2.

## Bioconductor 

Bioconductor is a repository of bioinformatics packages, which are not accessible from CRAN.
We can install bioconductor following [their instructions](http://bioconductor.org/install/). 
```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install()
```

This will install a set of core packages, additional packages (i.e maftools) can be installed 
by specifying their name.
```{r, eval=FALSE}
BiocManager::install("maftools")
``` 

We'll use bioconductor and maftools in [session 3](session3.Rmd)

## R markdown 

Compared to a script, which contains only your code and plain text comments, 
R markdown can be a very powerful way to present your R code, its output and comments in formatted text which can be shared in a wide range of formats including html pages, pdfs, word docs, and powerpoint presentations.  

The base idea is that there are code chunks, which get executed with the output printed below and formatted text. 
Behind the scene, this website is automatically generated by a series of R notebooks, available on [github](https://github.com/sdangelis/Bioinformatics-cafe-tutorials) but R Markdown can do much more!
You can read all about R Markdown in this [book](https://bookdown.org/yihui/rmarkdown/).

# Namespaces 

Functions are conventionally named with a short, descriptive verb.
Sometimes two packages use the same name for a function. In this case the most recently loaded package will overwrite the previous one
s
A very common case is the filter function in dplyr which overrides the default R function. Despite the shared name they do very different things. 

The base R filter function applies a linear filter to a time series
```{r eval=FALSE}
x <- 1:100
filter(x, rep(1, 3))
```
once we load dplyr the function filter filters rows in a data frame that satisfy a condition. In this case we can ask for hypothetical samples with less then 10 cells. Note: The variable cells is accessed directly rather than using `$cells`. This is one of the differences between normal and tidy R.

```{r, error=TRUE}
library(dplyr, verbose = F)

sample=1:10
cells=1:10 * 4
experimnent <- data.frame(sample, cells)
filter(experimnent, cells < 10) 
```

But what if we wanted the base R version? R has a very elegant built-in way to select which package we want to get our function from 
`namespace::function`^[Like C++ but unlike Python where the `import x as y` syntax is used to solve name conflicts].
Generally speaking all packages have their own names pace so we can explicitly ask for the filter function in base R ^[Built-in functions can belong to few different names paces,\n for instance base, stats, utils, graphics] with `stats::filter()`.
```{r}
x <- 1:100
stats::filter(x, rep(1, 3))
```

Arguably, it can be worth to use the `::` operator when you use a function from a package to explicitly state what packages are required and isolate that code from whatever package we have loaded with library.
This is suggested by the [Google R Style Guide](https://google.github.io/styleguide/Rguide.html).^[This introduces an additional nuance, calling library() will load said package in the global environment.\n affecting all your code, using :: will only affect that function call.\n :: is the to go when writing a package]


# Packages for session 2 

before the next session make sure you install the following packages 
- tidyverse
- cowplot
- Bioconductor

```{r, eval = FALSE, }
# you can use c() to create a vector of all the pacakges 
install.packages(c("tidyverse", "ggplot2", "cowplot"))
```